---
title: "Tumor Microenvironment"
author: "Ramon Massoni-Badosa"
date: "04/09/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

The objective of this notebook is to compare the composition and expression non-B cell peripheral blood cells between healthy, CLL and Richter samples.


# Pre-processing

## Load packages

```{r}
library(Seurat)
library(harmony)
library(SeuratWrappers)
library(tidyverse)
```


## Load data

```{r}
microenv_richter <- readRDS("/Volumes/Massoni_external/B_cell_atlas/RICHTER/current/results/R_objects/peripheral_blood_microenvironment_richter_seurat.rds")
microenv_cll_benchmark <- readRDS("/Volumes/Massoni_external/B_cell_atlas/JULIA/current/2-CLL/results/R_objects/cll_seurat_annotated.rds")
microenv_pbmc_benchmark <- readRDS("/Volumes/Massoni_external/B_cell_atlas/JULIA/current/1-PBMC/results/R_objects/10X_pbmc_Seurat_clustered.RDS")
```


## Source script with function definition

```{r}
source("bin/utils.R")
```


# Compositional analysis

```{r}
# Subset to good-quality cells (sampling time <= 2h or 4ºC)
Idents(microenv_pbmc_benchmark) <- "condition"
conditions_interest <- c("0h", "2h", "8h", "24h_4C", "48h_4C")
microenv_pbmc_benchmark <- subset(
  microenv_pbmc_benchmark,
  idents = conditions_interest
)
table(microenv_pbmc_benchmark$condition)
cll_selection <- microenv_cll_benchmark$temperature == "4C" | microenv_cll_benchmark$time %in% c("0h", "2h")
cells_cll <- colnames(microenv_cll_benchmark)[cll_selection]
microenv_cll_benchmark <- subset(microenv_cll_benchmark,cells = cells_cll)
table(microenv_cll_benchmark$temperature, microenv_cll_benchmark$time)


# Subset to peripheral blood cells and exclude B-cell lineage
Idents(microenv_pbmc_benchmark) <- "cell_type"
microenv_pbmc_benchmark <- subset(
  microenv_pbmc_benchmark,
  idents = c("T", "NK", "Monocyte")
)
Idents(microenv_cll_benchmark) <- "cell_type"
microenv_cll_benchmark <- subset(
  microenv_cll_benchmark,
  idents = c("T and NK", "Monocyte")
)


# Homogenize metadata and merge objects
str(microenv_pbmc_benchmark@meta.data)
str(microenv_cll_benchmark@meta.data)
str(microenv_richter@meta.data)
selected_cols_pbmc <- c("nCount_RNA", "nFeature_RNA", "donor", "cell_type")
microenv_pbmc_benchmark@meta.data <- microenv_pbmc_benchmark@meta.data[, selected_cols_pbmc]
new_colnames <- c("nCount_RNA", "nFeature_RNA", "donor", "original_annotation")
colnames(microenv_pbmc_benchmark@meta.data) <- new_colnames
microenv_cll_benchmark@meta.data <- microenv_cll_benchmark@meta.data[, selected_cols_pbmc]
colnames(microenv_cll_benchmark@meta.data) <- new_colnames
selected_cols_richter <- c("nCount_RNA", "nFeature_RNA", "patient_id", "cell_type", "is_richter")
microenv_richter@meta.data <- microenv_richter@meta.data[, selected_cols_richter]
new_colnames <- c("nCount_RNA", "nFeature_RNA", "donor", "original_annotation", "status")
colnames(microenv_richter@meta.data) <- new_colnames
microenv_pbmc_benchmark$status <- "healthy"
microenv_cll_benchmark$status <- "cll"
microenv_richter$status <- ifelse(microenv_richter$status == "richter", "richter", "cll")
microenv <- merge(
  microenv_pbmc_benchmark,
  c(microenv_cll_benchmark, microenv_richter)
)
```

Let us annotate the merged Seurat object:

```{r}
microenv <- NormalizeData(
  microenv,
  normalization.method = "LogNormalize",
  scale.factor = 1e4
)
microenv <- process_seurat_harmony(
  microenv,
  ndims_pca = 30,
  group_var = "donor", 
  ndims_harmony = 30
)
DimPlot(microenv, group.by = "status")


# Cluster
microenv <- FindNeighbors(microenv, reduction = "harmony", dims = 1:30)
microenv <- FindClusters(microenv, resolution = 0.4)
DimPlot(microenv)


# Markers 
markers_microenv <- FindAllMarkers(
  microenv, 
  only.pos = TRUE,
  test.use = "wilcox",
  verbose = TRUE,
  logfc.threshold = 0.5
)
markers_microenv <- markers_microenv %>%
  filter(p_val_adj < 1e-5) %>%
  arrange(cluster, desc(avg_logFC))
DT::datatable(markers_microenv)
```

| Cluster ID  | Markers    | Cell type              |   
|-------------|------------|------------------------|
| 0           | CD8A, CD8B | CD8+ T                 |
| 1           | GNLY, TRDC | NK/TRDC+ T             |
| 2           | IL7R, IL32 | Memory CD4+ T          |
| 3           | IL7R, CCR7 | Naive CD4+ T           |
| 4           | LYZ        | CD14+ Mono             |
| 5           | MT genes   | poor-quality           |
| 6           | NR4A2, CD8A| Exhausted CD8+ T       |
| 7           | MS4A7      | FCGR3A+ Mono           |
| 8           | CTLA4      | CTLA4+ CD4+ T          |
| 9           | IGHM       | doublets/misclassified |
| 10          | CD79A      | B                      |
| 11          | PPBP       | Platelet               |
| 12          | HBM        | Erythro                |
| 13          | FCER1A     | DC                     |


```{r}
microenv$new_annotation <- microenv$seurat_clusters
annotation_levels <- c("CD8+ T", "NK/TRDC+ T", "Memory CD4+ T", "Naive CD4+ T",
                       "CD14+ Mono", "poor-quality", "Exhausted CD8+ T",
                       "FCGR3A+ Mono", "CTLA4+ CD4+ T", "doublets/misclassified",
                       "B", "Platelet", "Erythro", "DC")
levels(microenv$new_annotation) <- annotation_levels


# Subset to eliminate poor-quality cells or cells too infrequent to establish conclusions
Idents(microenv) <- "new_annotation"
selected_types <- c("CD8+ T", "NK/TRDC+ T", "Memory CD4+ T", "Naive CD4+ T",
                    "CD14+ Mono", "Exhausted CD8+ T", "FCGR3A+ Mono",
                    "CTLA4+ CD4+ T")
microenv <- subset(microenv, idents = selected_types)
```

We noticed that there is one cluster that has NK markers and the expression of TCR delta (TCRD). Thus, this cluster can be further subdivided into two:

```{r}
FeaturePlot(microenv, c("GNLY", "TRDC"))
nk_tcrd <- subset(microenv, idents = "NK/TRDC+ T")
nk_tcrd <- process_seurat_harmony(
  nk_tcrd,
  ndims_pca = 30,
  group_var = "donor",
  ndims_harmony = 30
)
nk_tcrd <- FindNeighbors(nk_tcrd, reduction = "harmony", dims = 1:30)
nk_tcrd <- FindClusters(nk_tcrd, resolution = 0.2)
DimPlot(nk_tcrd)
markers_nk_tcrd <- FindAllMarkers(
  nk_tcrd, 
  only.pos = TRUE,
  test.use = "wilcox",
  verbose = TRUE,
  logfc.threshold = 0.3
)
markers_nk_tcrd <- markers_nk_tcrd %>%
  filter(p_val_adj < 1e-5) %>%
  arrange(cluster, desc(avg_logFC))
DT::datatable(markers_nk_tcrd)
```

| Cluster ID  | Markers     | Cell type             |   
|-------------|-------------|-----------------------|
| 0           | PFN1, KLRF1 | NK                    |
| 1           | TRDC        | TRDC+ T               |
| 2           | CIRBP, RBM3 | cold-shocked          |
| 3           | IL7R, LTB   | Misclassified         |  


```{r}
nk_cells <- colnames(nk_tcrd)[nk_tcrd$seurat_clusters == "0"]
trdc_cells <- colnames(nk_tcrd)[nk_tcrd$seurat_clusters == "1"]
microenv$new_annotation <- as.character(microenv$new_annotation)
microenv$new_annotation[colnames(microenv) %in% nk_cells] <- "NK"
microenv$new_annotation[colnames(microenv) %in% trdc_cells] <- "TRDC+ T"
Idents(microenv) <- "new_annotation"
selected_cells <- colnames(microenv)[microenv$new_annotation != "NK/TRDC+ T"]
microenv <- subset(microenv, cells = selected_cells)

# Reprocess
microenv <- process_seurat_harmony(
  microenv,
  ndims_pca = 30,
  group_var = "donor",
  ndims_harmony = 30
)
FeaturePlot(microenv, c("GNLY", "TRDC"))
DimPlot(microenv, label = TRUE) + NoLegend()
```

Finally, we can plot the stacked bar plot

```{r}
stacked_bar_df <- microenv@meta.data %>%
  group_by(new_annotation, status) %>%
  summarise(n_cells = n())
stacked_bar_df <- stacked_bar_df %>%
  group_by(status) %>%
  mutate(n_cells_total = sum(n_cells)) %>%
  ungroup() %>%
  mutate(pct_cells = n_cells / n_cells_total * 100)
stacked_bar_gg <- stacked_bar_df %>%
  ggplot(aes(status, pct_cells, fill = new_annotation)) +
    geom_col() +
    labs(x = "", y = "Percentage of Cells (%)", fill = "") +
    theme_classic() +
    theme(axis.title.y = element_text(size = 13)) +
    theme(axis.text.x = element_text(size = 11))
```


# Differential Expression Analysis

```{r}

```

